Compositional Verification of Concurrent C Programs with Search Structure Templates

### 2023-12-08 concurrency/CPP 2024/paper.tex
working on camera-ready

%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigplan,10pt, screen]{acmart}
\settopmatter{}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
%\AtBeginDocument{%
%  \providecommand\BibTeX{{%
%    Bib\TeX}}}
\startPage{1}
\bibliographystyle{ACM-Reference-Format}

%\usepackage{txfonts}
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
%% http://ctan.org/pkg/subcaption
\usepackage{uri}
\usepackage{mathtools}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{mathpartir}
\usepackage{semantic}
\usepackage{graphicx}
\usepackage{cases}
\usepackage{hyperref}
\usepackage{stmaryrd}
\usepackage{listings}
\usepackage{parcolumns}
\usepackage{scalerel}
%\usepackage{lstlangcoq}
\usepackage[edges]{forest}
\renewcommand{\lstlistingname}{Figure}

@@

@@

\section{Introduction}
\label{sec:introduction}
Krishna et al. ~\cite{templates} proposed concurrent search structure templates as a method for proving the correctness of concurrent data structures \emph{compositionally}, separating the proof of a concurrent access method (e.g., optimistic concurrency, hand-over-hand locking, forwarding via internal links) from the proof of the underlying data structure (e.g., linked list, hashtable, B-tree). The concurrency ``templates'' are verified parametrically over data structure operations, and the data structure operations are verified without any reference to concurrency. In theory, this could allow us to prove the correctness of $n$ concurrency patterns and $m$ (single-threaded) data structure implementations, and immediately obtain $n \times m$ verified concurrent data structures. In practice, the story is more complicated: certain patterns work only for specific data structures or require the data structures to store extra information, while some internal data structure operations may not fit the template model. %\wm{Maybe say more about the idea that they only built a proof of concept, and there are still important details not worked out/only tested on ad-hoc examples.}

@@

Our specific contributions are:
\begin{itemize}
	\item We reimplement the template approach independently of flow interfaces, with a simple interface involving only the concept of ``keys belonging in this node/substructure''.
	\item We implement the template approach in VST, allowing us to apply it to C programs and obtain end-to-end correctness proofs in a single verification system.
	\item To the best of our knowledge, this is the first mechanized verification of a template approach to concurrent data structure implementations in a real-world programming language, and its first application to data structures not written specifically as case studies.
	\item We give a precise description of search structure templates, and identify places where it is difficult in practice to preserve the boundary between template and data structure.
\end{itemize}

\subsection*{Related Work}
Recent years have seen major advances in concurrent separation logics (CSLs) for verifying fine-grained concurrent programs, including Iris~\cite{iris}, VST~\cite{plcc,iris-vst-arxiv}, FCSL~\cite{fcsl}, TaDA~\cite{tada}, and VeriFast~\cite{verifast-conc}. The innovations of Iris (custom ghost state) and TaDA (logical atomicity) were particularly essential for the formalization of the search structure template approach. Some of the more complex data structures verified include Java's \lstinline{ConcurrentSkipListMap}~\cite{Xiong2017Abstract} and a multi-producer multi-consumer concurrent queue from the Folly library~\cite{iris-folly}. Notably, VST is the only one of these systems that is all three of 1) mechanized (i.e., implemented in a theorem prover), 2) foundational (i.e., connected to a formal semantics of the target language), and 3) targeting an implementation language (C) rather than a toy language or algorithmic representation. %do we want to say this part?

@@

\begin{figure}[h]
	\begin{subfigure}[t]{0.45\textwidth}
		\lstinputlisting[language=caml, style=myStyle]{lock_traverse.ml} 
		%		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in an ML-like language \cite{krishna2019compositional}} 
		%		\label{traverse_lock_a}	
	\end{subfigure}
	\qquad
	\begin{subfigure}[t]{0.45\textwidth}
		\lstinputlisting[language=caml, style=myStyle]{insert.ml} 
	\end{subfigure}
		%		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in C}
		%		\label{traverse_lock_b}
	\caption{The lock-coupling search structure template}
	\label{template-ex}
\end{figure}

@@

\subsubsection{Logical atomicity} 
%In this section, we show how to verify the templates presented above, by proving that each \lstinline{traverse} function meets its specification. %The detailed exploration of these template verifications will be discussed in Section \ref{traverse_proof_lock} and \ref{traverse_proof_giveup}.

@@

As an example, a sequential stack \lstinline{push} operation could conventionally be specified as 
$$
 {\color{specblue} \{\mathsf{stack}\ s\ p\} } \ \texttt{push}(p, v)\  {\color{specblue} \{\mathsf{stack}\ (v :: s)\ p\}}$$  Its concurrent counterpart could be specified as 
\begin{mathpar}
	{\color{specblue} \fforall s.\ \left\langle \mathsf{is\_stack}_g\ s\ |\ \mathsf{stack}_g\ vs\right\rangle\ } 
	 \vspace{-0.8em}  \\ \texttt{push}(s, v)\   \vspace{-0.8em}  \\ 
	{\color{specblue} \left\langle \mathsf{is\_stack}_g\ s\ |\ \mathsf{stack}_g\ (v::vs)\right\rangle }
\end{mathpar}
indicating that the \lstinline{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transforming the stack from $vs$ to $v::vs$ at some point during its execution. The stack itself is a shared resource, and can only be accessed and modified atomically by threads holding the corresponding \texttt{is\_stack} assertion. The local and shared assertions are connected by an arbitrary identifier $g$, which we will generally omit when clear from context. Logical atomicity can be used to prove linearizability---if all of a data structure's operations satisfy logically atomic triples derived from the corresponding sequential operations, the data structure is linearizable~\cite{la-lin}---but can also be used to specify more complex behavior of nonlinearizable data structures~\cite{compass}.

@@

A concurrent search structure is a data structure that supports three operations: \lstinline{insert}, \lstinline{lookup}, and \lstinline{delete}. It is intended to implement a map from keys to values, and to behave correctly when accessed simultaneously by any number of threads. In separation logic, we can prove correctness of a search structure by showing that its operations satisfy the following logically atomic specifications, where $\nodeboxrep$ is a per-thread handle to the data structure and $\treerep$ is a shared assertion linking the values in memory to an abstract map $m$ from keys to values:
\begin{mathpar}
	{\color{specblue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m)
		\right\rangle
	} \vspace{-0.85em} \\ \texttt{insert(r, k, v)}\ \vspace{-0.85em} \\
	{\color{specblue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle
	}
\\
	{\color{specblue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m)
		\right\rangle
	} \vspace{-0.8em}
	\\ \texttt{lookup(r, k)}\  \vspace{-0.8em} \\
	{\color{specblue}
		\left\langle \texttt{v}. \ \
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m) \land m(\texttt{k}) = \texttt{v}
		\right\rangle
	}
	\\
	{\color{specblue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m)
		\right\rangle
	}  \vspace{-0.8em} 
	\\ \texttt{delete(r, k)}\ \vspace{-0.8em}\\
	{\color{specblue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \_ ])
		\right\rangle
	}
\end{mathpar}
In other words, each operation takes effect atomically on the abstract map $m$, reading and/or updating it as appropriate to the operation. Krishna et al. use flow interfaces~\cite{krishna2017flow} to implement the $\treerep$ assertion connecting the data structure implementation to the abstract map, but this is not fundamental to the approach: any definition of $\nodeboxrep$ and $\treerep$ that can be used to prove the specifications (and is nontrivial, i.e., holds on the initial state of the data structure) will yield a correct data structure.

A search structure template is an implementation of these three functions, written in a particularly restricted way. Aside from some top-level management code (e.g., acquiring and releasing an initial lock), all that each function should do is call two functions: a function called \lstinline{traverse}, which is specific to the template and implements its synchronization mechanism, and a concurrency-unaware function (e.g. \lstinline{insertOp}) provided by the target data structure. (In fact, Krishna et al. implemented all three operations as a single function \lstinline{decisiveOp}, parameterized by a function that takes the target operation type as input and performs the appropriate data structure operation; for ease of adaptation to C, we write three different functions, but their structures are identical.) A template consists of an implementation of the top-level operations and the \lstinline{traverse} function used to define them; a data structure consists of an implementation of \lstinline{insertOp}, etc., as well as a function \lstinline{findNext} that is used to implement \lstinline{traverse}. The interface between them is a set of standard specifications for \lstinline{insertOp}, etc., and \lstinline{findNext}: if a template's functions can be verified assuming these standard specifications, and a data structure's implementations can be shown to meet those specifications, then the template and the data structure can be combined to yield a verified concurrent search structure.

@@

In other words, $\inFP(n)$ is sufficient to guarantee that node $n$ is in the abstract state of the data structure and its lock protects associated resources $\mathsf{R}(n)$, the \emph{lock invariant} for the node. 
%In this case, $\mathsf{R}$ contains the contents of the node (i.e., the \lstinline{node} struct contained inside the \lstinline{node_t} struct of \lstinline{n}).

%\wm{The fonts are still off here---for instance, p, n, and n' are mathematical variables, not references to the C code. Fix if we have time.}
The \lstinline{traverse} function can then be specified as follows:
%\than{Try to avoid using specific name of $\mathrm{tree\_rep}$ for public condition, I suggested to use $\mathrm{Node}$ for generic purpose.} 
\begin{mathpar}
{ \color{specblue}
	\fforall \  m. \ \scaleleftright[1.4ex]{\langle}
	{\begin{array}{l}
			\texttt{pn} \mapsto (p, n) \ \ast \ 
			\inFP(n)  \ast \ \mathsf{R}(n) 
		\end{array}  
		\Big \vert  \treerep\ (m)
	}  
	{\rangle}
}
\vspace{-6pt}\\
\vspace{-6pt}\texttt{traverse(pn, k)} \\
{ \color{specblue}
 \scaleleftright[1.4ex]{\langle}
{\begin{array}{l}
		\mathit{res}. \exists \  n', v, \mathsf{range}. \\ 
		\ \ \ \ \ \  \texttt{pn} \mapsto \left(n', n'\right) \ast  \inFP\left(n'\right) \ast\ \texttt{k} \in \mathsf{range}\ \ast \ 
		\\ 
		\ \ \ \ \ \ \mathsf{if} \ \mathit{res}  \ \mathsf{then}  \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
		\\ 
		\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
\end{array}  
\Bigg \vert \treerep\ (m)
}  
{\rangle}
}
\end{mathpar}

The local precondition of \lstinline{traverse} includes both the node handle $\inFP(n)$ and its contents $\mathsf{R}(n)$, indicating that a thread should already hold $n$'s lock before calling \lstinline{traverse}. The output of \lstinline{traverse} is a new node $n'$ in the data structure such that the key \texttt{k} falls within the range of $n'$. The local postcondition then includes the handle of the new node $\inFP(n')$, its contents $\mathsf{R}(n')$, and a Boolean variable $\mathit{res}$ indicating whether \lstinline{traverse} found an empty node or a node with key \lstinline{k}.

The resources $\mathsf{R}$ contained in a node depend on the specific data structure, but always include a piece of \emph{ghost state} describing the current state of the node (its key, value, and key range) shared between the invariant $\mathsf{R}$ and the abstract state $\treerep$, ensuring that the lock and the abstract state agree on the contents of the node. Formally, the lock invariant is defined by 
\begin{align*}&\mathsf{node\_contents}(n, c, \mathsf{range}) \triangleq \\ 
	&\qquad\mathsf{ghost\_node}(n, c,\mathsf{range}) \ast \mathsf{node\_data}(n, c) \\
	&\mathsf{R}(n) \triangleq \exists\ c, \mathsf{range}.\ \mathsf{node\_contents}(n, c, \mathsf{range})
\end{align*}
where the definition of $\mathsf{node\_data}$ is supplied by the target data structure. The contents $c$ of a node can be either a key-value pair $(k, v)$, or the empty contents $\cdot$ (used for nodes that have been allocated but not yet assigned keys). Then $\treerep\ (m)$ is defined as a collection of $\mathsf{ghost\_node}$s that form a tree containing all the key-value pairs in $m$.

@@

The key to the correctness of the \texttt{traverse} function is the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}, \texttt{k}) \triangleq \\ \exists \ p, n, c, \mathsf{range}.\ &\texttt{pn} \mapsto (p, n) \ast \texttt{k} \in \mathsf{range}\ \ast \ \\  \inFP(n)  \ \ast \  &\mathsf{node\_contents}(n, c, \mathsf{range})
\end{align*}

\begin{figure*}[!ht]
	$\color{specblue}
	\fforall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
	\inFP(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ (m) \
	\right\rangle$
		\begin{minipage}{0.9\textwidth}
			\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_traverse.c}
		\end{minipage}
	$\color{specblue}
	\left\langle \mathit{res}. \ \exists \  n', v.
	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \inFP(n') \ \ast\ \texttt{k} \in \mathsf{range}\ \ast\ 
		\\ 
		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
	\end{array}
	\ \Bigg| \treerep\ (m) \
	\right\rangle$
	\caption{Proof outline of the lock-coupling \texttt{traverse} function}
	\label{proof_lock_traverse}
\end{figure*}

@@

\subsubsection{Verifying the Give-Up Template}
\label{traverse_proof_giveup}
The give-up template's \texttt{traverse} specification is almost the same as in the lock-coupling template, except that the caller does not need to hold any locks before calling it, so the invariant $\mathsf{R}$ does not appear in the precondition:
\begin{mathpar}
	{ \color{specblue}
		\fforall \  m. \ \scaleleftright[1.4ex]{\langle}
		{\begin{array}{l}
				\texttt{pn} \mapsto (p, n) \ast  
				\inFP(n)   
			\end{array}  
			\Big \vert  \treerep\ (m)
		}  
		{\rangle}
	}
	\vspace{-6pt}\\
	\vspace{-6pt}\texttt{traverse(pn, k)} \\
	{ \color{specblue}
		\scaleleftright[1.4ex]{\langle}
		{\begin{array}{l}
				\mathit{res}. \exists \  n', v, \mathsf{range}. \\ 
				\ \ \ \ \ \  \texttt{pn} \mapsto \left(n', n'\right) \ast  \inFP\left(n'\right) \ast \texttt{k} \in \mathsf{range} \ \ast 
				\\ 
				\ \ \ \ \ \ \mathsf{if} \ \mathit{res}  \ \mathsf{then}  \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
				\\ 
				\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
			\end{array}  
			\Bigg \vert \treerep\ (m)
		}  
		{\rangle}
	}
\end{mathpar}

>>>>

%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigplan,10pt, screen]{acmart}
\settopmatter{}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
%\AtBeginDocument{%
%  \providecommand\BibTeX{{%
%    Bib\TeX}}}
\startPage{1}

%\usepackage{txfonts}
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
%% http://ctan.org/pkg/subcaption
\usepackage{uri}
\usepackage{mathtools}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{tikzit}
\input{box.tikzstyles}
\usetikzlibrary{fit}
\usepackage{mathpartir}
\usepackage{semantic}
\usepackage{graphicx}
\usepackage{cases}
\usepackage{hyperref}
\usepackage{stmaryrd}
\usepackage{listings}
\usepackage{parcolumns}
\usepackage{scalerel}
%\usepackage{lstlangcoq}
\usepackage[edges]{forest}
\renewcommand{\lstlistingname}{Figure}

@@

%https://tex.stackexchange.com/questions/112634/how-to-draw-a-diagram-with-boxes-of-text-inside-each-other-and-have-them-laid-ou
\newcommand\TBox[2][]{%
  \tikz\node[draw,align=left,#1] {#2};\hskip2pt}

\usetikzlibrary{positioning}

@@

\section{Introduction}
\label{sec:introduction}
Concurrent search structure templates~\cite{shasha1988concurrent} are a technique for describing and implementing concurrency patterns (e.g., optimistic concurrency, hand-over-hand locking, forwarding via internal links) that can then be specialized to various search data structures (e.g., linked list, hashtable, B-tree) compositionally. Krishna et al. ~\cite{templates} repurposed templates as a compositional \emph{verification} technique, separating the proof of a concurrent access method from the proof of the underlying data structure. The concurrency templates are verified parametrically over data structure operations, and the data structure operations are verified without any reference to concurrency. In theory, this could allow us to prove the correctness of $n$ concurrency patterns and $m$ (single-threaded) data structure implementations, and immediately obtain $n \times m$ verified concurrent data structures. In practice, the story is more complicated: certain patterns work only for specific data structures or require the data structures to store extra information, while some internal data structure operations may not fit the template model. %\wm{Maybe say more about the idea that they only built a proof of concept, and there are still important details not worked out/only tested on ad-hoc examples.}

@@

Our specific contributions are:
\begin{itemize}
	\item We reimplement the template approach independently of flow interfaces, with a simple interface involving only the concept of ``keys belonging in this node/substructure''.
	\item We implement the template approach in VST, allowing us to apply it to C programs and obtain end-to-end correctness proofs in a single verification system.
	\item To the best of our knowledge, this is the first mechanized verification of a template approach to concurrent data structure implementations in a real-world programming language, and its first application to data structures not written specifically as case studies.
	\item We give a precise description of search structure templates, and identify places where it is difficult in practice to preserve the boundary between template and data structure: for instance, both node creation and rotation in binary search trees operate on concurrent and data-structure-specific aspects in ways that do not cleanly decompose.
\end{itemize}

\subsection*{Related Work}
Recent years have seen major advances in concurrent separation logics (CSLs) for verifying fine-grained concurrent programs, including Iris~\cite{iris}, VST~\cite{plcc,iris-vst-arxiv}, FCSL~\cite{fcsl}, TaDA~\cite{tada}, and VeriFast~\cite{verifast-conc}. The innovations of Iris (custom ghost state) and TaDA (logical atomicity) were particularly essential for the formalization of the search structure template approach. Some of the more complex data structures verified include Java's \lstinline{ConcurrentSkipListMap}~\cite{Xiong2017Abstract} and a multi-producer multi-consumer concurrent queue from the Folly library~\cite{iris-folly}. Notably, VST is the only one of these systems that is all three of 1) mechanized (i.e., implemented in a theorem prover), 2) foundational (i.e., connected to a formal semantics of the target language), and 3) targeting an implementation language (C) rather than a toy language or algorithmic representation. For instance, Iris's RustBelt instance~\cite{rustbelt} targets a core calculus based on Rust's intermediate language and has been used to verify several interesting concurrent programs, but does not provide guarantees about actual running Rust code.

@@

\begin{figure}[h]
	\begin{subfigure}[t]{0.45\textwidth}
		\lstinputlisting[language=caml, style=myStyle, numbers=none]{lock_traverse.ml} 
		%		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in an ML-like language \cite{krishna2019compositional}} 
		%		\label{traverse_lock_a}	
	\end{subfigure}
	\qquad
	\begin{subfigure}[t]{0.45\textwidth}
		\lstinputlisting[language=caml, style=myStyle, numbers=none]{insert.ml} 
	\end{subfigure}
		%		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in C}
		%		\label{traverse_lock_b}
	\caption{The lock-coupling search structure template}
	\label{template-ex}
\end{figure}

@@

\subsubsection{Logical atomicity} 
\label{logatom}
%In this section, we show how to verify the templates presented above, by proving that each \lstinline{traverse} function meets its specification. %The detailed exploration of these template verifications will be discussed in Section \ref{traverse_proof_lock} and \ref{traverse_proof_giveup}.

@@

As an example, a sequential stack \lstinline{push} operation could conventionally be specified as 
$$
 {\color{specblue} \{\mathsf{stack}(s, p)\} } \ \texttt{push}(p, v)\  {\color{specblue} \{\mathsf{stack}(v :: s, p)\}}$$  Its concurrent counterpart is canonically specified as 
\begin{mathpar}
	{\color{specblue} \fforall s.\ \left\langle \mathsf{is\_stack}_g(p)\ |\ \mathsf{stack}_g(s)\right\rangle\ } 
	 \vspace{-0.8em}  \\ \texttt{push}(p, v)\   \vspace{-0.8em}  \\ 
	{\color{specblue} \left\langle \mathsf{is\_stack}_g(p)\ |\ \mathsf{stack}_g(v::s)\right\rangle }
\end{mathpar}
indicating that the \lstinline{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transforming the stack from $vs$ to $v::vs$ at some point during its execution. The stack itself is a shared resource, and can only be accessed and modified atomically by threads holding the corresponding \texttt{is\_stack} assertion. The local and shared assertions are connected by an arbitrary identifier $g$, which we will generally omit when clear from context. Logical atomicity can be used to prove linearizability, the standard correctness condition for concurrent data structures---if all of a data structure's operations satisfy logically atomic triples derived from the corresponding sequential operations, the data structure is linearizable~\cite{la-lin}. It can also be used to specify more complex behavior of nonlinearizable data structures~\cite{compass}, but in this paper we will focus on linearizable data structures and canonical specifications.

@@

A concurrent search structure is a data structure that supports three operations: \lstinline{insert}, \lstinline{lookup}, and \lstinline{delete}. It is intended to implement a map from keys to values, and to behave correctly when accessed simultaneously by any number of threads. In separation logic, we can prove correctness of a search structure by showing that its operations satisfy the following logically atomic specifications, where $\nodeboxrep$ is a per-thread handle to the data structure and $\treerep$ is a shared assertion linking the values in memory to an abstract map $m$ from keys to values (we omit the ghost identifier $g$ connecting $\nodeboxrep$ and $\treerep$, as described in section~\ref{logatom}):
\begin{mathpar}
	{\color{specblue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m)
		\right\rangle
	} \vspace{-0.85em} \\ \texttt{insert(r, k, v)}\ \vspace{-0.85em} \\
	{\color{specblue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle
	}
\\
	{\color{specblue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m)
		\right\rangle
	} \vspace{-0.8em}
	\\ \texttt{lookup(r, k)}\  \vspace{-0.8em} \\
	{\color{specblue}
		\left\langle \texttt{v}. \ \
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m) \land m(\texttt{k}) = \texttt{v}
		\right\rangle
	}
	\\
	{\color{specblue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m)
		\right\rangle
	}  \vspace{-0.8em} 
	\\ \texttt{delete(r, k)}\ \vspace{-0.8em}\\
	{\color{specblue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \_ ])
		\right\rangle
	}
\end{mathpar}
In other words, each operation takes effect atomically on the abstract map $m$, reading and/or updating it as appropriate to the operation. Krishna et al. use flow interfaces~\cite{krishna2017flow} to implement the $\treerep$ assertion connecting the data structure implementation to the abstract map, but this is not fundamental to the approach: any definition of $\nodeboxrep$ and $\treerep$ that can be used to prove the specifications (and is nontrivial, i.e., holds on the initial state of the data structure) will yield a correct data structure.

\begin{figure}[htb]
\ctikzfig{templates}
	\caption{Diagram of the template approach. Each pair of template and data structure should yield a verified concurrent data structure.}
	\label{templates-diagram}
\end{figure}

A search structure template is an implementation of these three functions, written in a particularly restricted way outlined in Figure~\ref{templates-diagram}. Aside from some top-level management code (e.g., acquiring and releasing an initial lock), all that each function should do is call two functions: a helper function called \lstinline{traverse}, which is specific to the template and implements its synchronization mechanism, and a concurrency-unaware function (e.g. \lstinline{insertOp}) provided by the target data structure. (In fact, Krishna et al. implemented all three operations as a single function \lstinline{decisiveOp}, parameterized by a function that takes the target operation type as input and performs the appropriate data structure operation; for ease of adaptation to C, we write three different functions, but their structures are identical.) A template consists of an implementation of the top-level operations and the \lstinline{traverse} function used to define them; a data structure consists of an implementation of \lstinline{insertOp}, etc., as well as a function \lstinline{findNext} that is used to implement \lstinline{traverse}. The interface between them is a set of standard specifications for \lstinline{insertOp}, etc., and \lstinline{findNext}: if a template's functions can be verified assuming these standard specifications, and a data structure's implementations can be shown to meet those specifications, then the template and the data structure can be combined to yield a verified concurrent search structure.

@@

In other words, $\inFP(n)$ is sufficient to guarantee that node $n$ is in the abstract state of the data structure and its lock protects associated resources $\mathsf{R}(n)$, the \emph{lock invariant} for the node, which we will soon describe in detail. 
%In this case, $\mathsf{R}$ contains the contents of the node (i.e., the \lstinline{node} struct contained inside the \lstinline{node_t} struct of \lstinline{n}).

%\wm{The fonts are still off here---for instance, p, n, and n' are mathematical variables, not references to the C code. Fix if we have time.}
The \lstinline{traverse} function can then be specified as follows:
%\than{Try to avoid using specific name of $\mathrm{tree\_rep}$ for public condition, I suggested to use $\mathrm{Node}$ for generic purpose.} 
\begin{mathpar}
{ \color{specblue}
	\fforall \  m. \ \scaleleftright[1.4ex]{\langle}
	{\begin{array}{l}
			\texttt{pn} \mapsto (p, n) \ \ast \ 
			\inFP(n)  \ast \ \mathsf{R}(n) 
		\end{array}  
		\Big \vert  \ \treerep\ (m)
	}  
	{\rangle}
}
\vspace{-6pt}\\
\vspace{-6pt}\texttt{traverse(pn, k)} \\
{ \color{specblue}
 \scaleleftright[1.4ex]{\langle}
{\begin{array}{l}
		\mathit{res}.\ \exists \  n', v, \mathit{range}. \\ 
		\ \ \ \ \ \  \texttt{pn} \mapsto \left(n', n'\right) \ast  \inFP\left(n'\right) \ast\ \texttt{k} \in \mathit{range}\ \ast \ 
		\\ 
		\ \ \ \ \ \ \mathsf{if} \ \mathit{res}  \ \mathsf{then}  \ \mathsf{node\_contents}(n', \cdot, \mathit{range}) \ 
		\\ 
		\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathit{range})
\end{array}  
\Bigg \vert \ \treerep\ (m)
}  
{\rangle}
}
\end{mathpar}

The local precondition of \lstinline{traverse} includes both the node handle $\inFP(n)$ and its contents $\mathsf{R}(n)$, indicating that a thread should already hold $n$'s lock before calling \lstinline{traverse}. The output of \lstinline{traverse} is a new node $n'$ in the data structure such that the key \texttt{k} falls within the range of $n'$. The local postcondition then includes the handle of the new node $\inFP(n')$, its contents $\mathsf{node\_contents}(n', ...)$, and a Boolean variable $\mathit{res}$ indicating whether \lstinline{traverse} found an empty node or a node with key \lstinline{k} (which is reflected in the remaining arguments to $\mathsf{node\_contents}$).

The resources $\mathsf{R}$ contained in a node depend on the specific data structure, but always include a piece of \emph{ghost state} describing the current state of the node (its key, value, and key range) shared between the invariant $\mathsf{R}$ and the abstract state $\treerep$, ensuring that the lock and the abstract state agree on the contents of the node. Formally, the lock invariant is defined by 
\begin{align*}&\mathsf{node\_contents}(n, c, \mathit{range}) \triangleq \\ 
	&\qquad\mathsf{ghost\_node}(n, c,\mathit{range}) \ast \mathsf{node\_data}(n, c) \\
	&\mathsf{R}(n) \triangleq \exists\ c, \mathit{range}.\ \mathsf{node\_contents}(n, c, \mathit{range})
\end{align*}
where the definition of $\mathsf{node\_data}$ is supplied by the target data structure. The contents $c$ of a node can be either a key-value pair $(k, v)$, or the empty contents $\cdot$ (used for nodes that have been allocated but not yet assigned keys). Then $\treerep\ (m)$ is defined as a collection of $\mathsf{ghost\_node}$s that form a tree containing all the key-value pairs in $m$.

@@

The key to the correctness of the \texttt{traverse} function is the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}, \texttt{k}) \triangleq \\ \exists \ p, n, c, \mathit{range}.\ &\texttt{pn} \mapsto (p, n) \ast \texttt{k} \in \mathit{range}\ \ast \ \\  \inFP(n)  \ \ast \  &\mathsf{node\_contents}(n, c, \mathit{range})
\end{align*}

\begin{figure*}[!ht]
	$\color{specblue}
	\fforall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
	\inFP(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ (m) \
	\right\rangle$
		\begin{minipage}{0.9\textwidth}
			\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_traverse.c}
		\end{minipage}
	$\color{specblue}
	\left\langle \mathit{res}. \ \exists \  n', v.
	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \inFP(n') \ \ast\ \texttt{k} \in \mathit{range}\ \ast\ 
		\\ 
		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathit{range}) \ 
		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathit{range})
	\end{array}
	\ \Bigg| \ \treerep\ (m) \
	\right\rangle$
	\caption{Proof outline of the lock-coupling \texttt{traverse} function}
	\label{proof_lock_traverse}
\end{figure*}

@@

\subsubsection{Verifying the Give-Up Template}
\label{traverse_proof_giveup}
The give-up template's \texttt{traverse} specification is almost the same as in the lock-coupling template, except that the caller does not need to hold any locks before calling it, so the invariant $\mathsf{R}$ does not appear in the precondition:
\begin{mathpar}
	{ \color{specblue}
		\fforall \  m. \ \scaleleftright[1.4ex]{\langle}
		{\begin{array}{l}
				\texttt{pn} \mapsto (p, n) \ast  
				\inFP(n)   
			\end{array}  
			\Big \vert \ \treerep\ (m)
		}  
		{\rangle}
	}
	\vspace{-6pt}\\
	\vspace{-6pt}\texttt{traverse(pn, k)} \\
	{ \color{specblue}
		\scaleleftright[1.4ex]{\langle}
		{\begin{array}{l}
				\mathit{res}.\ \exists \  n', v, \mathit{range}. \\ 
				\ \ \ \ \ \  \texttt{pn} \mapsto \left(n', n'\right) \ast  \inFP\left(n'\right) \ast \texttt{k} \in \mathit{range} \ \ast 
				\\ 
				\ \ \ \ \ \ \mathsf{if} \ \mathit{res}  \ \mathsf{then}  \ \mathsf{node\_contents}(n', \cdot, \mathit{range}) \ 
				\\ 
				\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathit{range})
			\end{array}  
			\Bigg \vert \ \treerep\ (m)
		}  
		{\rangle}
	}
\end{mathpar}
